---
title: "BBS NextGen committee"
output: 
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    self_contained: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
---
```{r setup, include = FALSE, echo = FALSE}
## load packages
packs.html <- c("knitr", "pander", "dplyr", "readxl")
for (i in 1:length(packs.html)){library(packs.html[i], character.only = TRUE)}

knitr::opts_chunk$set(echo = TRUE)

## upcoming   
path <- getwd()
upcoming <- as.data.frame(read_excel(paste(path, "/data/upcoming.xlsx", sep = ""), col_types = NULL))

lunch <- upcoming |> filter(tolower(event) == "bbs nextgen lunch")
has_lunch <- nrow(lunch) > 0
```
# Activities

The BBS NextGen committee

* coordinates BBS next generation events (typically two each year, see the [upcoming](events_upcoming.html) and [past](events_past.html) events),
* coordinates the annual BBS next generation mentoring program which kicked off in February 2025 (see [the slide deck on the BBS next generation mentoring program](data/BBS NextGen_Mentoring_Program.pdf) for more detail),
* maintains the [BBS next generation mailing list](index.html#additional-e-mailing-list-of-the-bbs-next-generation-committee) and [LinkedIn page](https://www.linkedin.com/company/bbs-next-generation), as well as
* organizes monthly community / networking lunches (see the [LinkedIn posts](https://www.linkedin.com/company/bbs-next-generation/posts) for upcoming ones).  
  `{r} I(ifelse(has_lunch, paste0("The next lunch will be at KLARA (Clarastrasse 13, 4058 Basel) from 12-13 CET on ", lunch$date, " - [RSVP here](", lunch$registration, ")."), ""))`

For a detailed overview, please see the following slides for the [2024 BBS Next Gen "Today, Tomorrow and the Future"](data/BBS NextGen_Today_Tomorrow_Future_2024.pdf).

# Committee members

```{r nextgen_board, include = FALSE, echo = FALSE}
## roster   
path <- getwd()
roster <- read_excel("data/roster_nextgen.xlsx", na = character())
roster$Name <- paste(roster$First, roster$Last, sep = " ")
roster <- roster |>
  filter(Committee == 1) |>
  arrange(Company, Last, First) |>
  select(Name, Company) 

# roster[is.na(roster)] <- ""
```

```{r, include=TRUE, echo=FALSE}
kable(roster)
```



