---
title: "BBS NextGen committee"
output: 
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    self_contained: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
---

# Activities

The BBS NextGen committee

* coordinates BBS next generation events (typically two each year, see the [upcoming](events_upcoming.html) and [past](events_past.html) events),
* maintains the [BBS next generation mailing list](index.html#additional-e-mailing-list-of-the-bbs-next-generation-committee) and [LinkedIn page](https://www.linkedin.com/company/bbs-next-generation), as well as
* organizes monthly community / networking lunches (see the [LinkedIn posts](https://www.linkedin.com/company/bbs-next-generation/posts) for upcoming ones).

# Committee members

```{r setup, include = FALSE, echo = FALSE}
## load packages
packs.html <- c("knitr", "pander", "dplyr", "readxl")
for (i in 1:length(packs.html)){library(packs.html[i], character.only = TRUE)}

# knitr::opts_chunk$set(echo = TRUE)

## roster   
path <- getwd()
roster <- read_excel("data/roster_nextgen.xlsx", na = character())
roster$Name <- paste(roster$First, roster$Last, sep = " ")
roster <- roster |>
  filter(Committee == 1) |>
  arrange(Company, Last, First) |>
  select(Name, Company) 

# roster[is.na(roster)] <- ""
```

```{r, include=TRUE, echo=FALSE}
kable(roster)
```



