---
title: "Advanced group-sequential and adaptive confirmatory clinical trial designs, with R practicals using rpact"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: Contents
    toc-location: left
    number-sections: true
    number-depth: 2
--- 

# Purpose of this document

Materials for the [CEN 2023](https://cen2023.github.io/home/index.html) pre-conference course [Advanced group-sequential and adaptive confirmatory clinical trial designs, with R practicals using rpact](https://cen2023.github.io/home/courses.html#Advanced_group-sequential_and_adaptive_confirmatory_clinical_trial_designs,_with_R_practicals_using_rpact), held in Basel on 3rd September 2023 by [Kaspar Rufibach](mailto:kaspar.rufibach@roche.com), [Marc Vandemeulebroecke](mailto:marc.vandemeulebroecke@novartis.com), [Gernot Wassmer](mailto:gernot.wassmer@rpact.com), and [Marcel Wolbers](mailto:marcel.wolbers@roche.com). 

# Agenda of the event

| 09:00-09:10: Welcome                                                                  	
| 09:10-09:50: Theory: Efficient use of interim analyses (Kaspar)                       	
| 09:50-10:30: Introduction to rpact, part 1 (Gernot), exercises (Kaspar & all, ex 1)   	
| 10:30-11:00: Coffee break                                                             	
| 11:00-11:40: Theory: Introduction adaptive trials & sample size re-calculation (Marc) 	
| 11:40-12:30: Introduction rpact, part 2 (Gernot), exercises (Marc & all, ex 2)      	
| 12:30-14:00: Lunch                                                                    	
| 14:00-15:30: Theory: Multi-arm multi-stage designs (Marcel & Kaspar)                  	
| 15:30-16:00: Coffee break                                                             	
| 16:00-17:00: Introduction rpact, part 3 (Gernot), exercises (Gernot & all, ex 3-5)      	
| 17:00-17:15: Teaser: What can we expect from ICH E20? (Uli & Frank)                   	
| 17:15-17:30: Final Q&A                                                                	                                 	

# Preparation prior to the event

- Please download all the documents from this webpage: [link to zip file](BBS_adaptive.zip).
- Install the [latest version of rpact](https://cran.r-project.org/package=rpact) locally on your computer:

```{r, include=TRUE, echo=TRUE, warning=FALSE, eval = FALSE}
install.packages("rpact")
packageVersion("rpact")
```

# Documents

```{r, include=TRUE, echo=FALSE, message = FALSE}
suppressWarnings(library(knitr))

tab <- data.frame(matrix("", ncol = 4, nrow = 1),stringsAsFactors = FALSE)
colnames(tab) <- c("Presenter", "Title", "Slides", "Markdown accompanying slides")
tab[1, "Presenter"]                    <- "Kaspar Rufibach"
tab[1, "Title"]                        <- "Efficient use of futility and efficacy interim analyses in group-sequential designs" 
tab[1, "Slides"]                       <- "[X](RufibachWolbers_efficient_interims_handout.pdf)"
tab[1, "Markdown accompanying slides"] <- "[X](RufibachWolbers_efficient_interims.html)"

tab[2, "Presenter"]                    <- "Marc Vandemeulebroecke"
tab[2, "Title"]                        <- "Introduction adaptive trials & sample size re-calculation" 
tab[2, "Slides"]                       <- "[X](Vandemeulebroecke_intro_adaptive.pdf)"
tab[2, "Markdown accompanying slides"] <- ""

tab[3, "Presenter"]                    <- "Gernot Wassmer"
tab[3, "Title"]                        <- "Introduction to rpact" 
tab[3, "Slides"]                       <- "[X](Wassmer_intro_rpact.pdf)"
tab[3, "Markdown accompanying slides"] <- ""

tab[4, "Presenter"]                    <- "Marcel Wolbers & Kaspar Rufibach"
tab[4, "Title"]                        <- "Multi-arm multi-stage designs (MAMS)" 
tab[4, "Slides"]                       <- "[X](CEN2023_MAMSsection_Wolbers_Rufibach_handout.pdf)"
tab[4, "Markdown accompanying slides"] <- ""

tab[5, "Presenter"]                    <- "Frank Bretz and Uli Burger"
tab[5, "Title"]                        <- "Adaptive trials: Some general considerations" 
tab[5, "Slides"]                       <- ""
tab[5, "Markdown accompanying slides"] <- ""

kable(tab, align = "llcc")
```

* Exercises: [markdown as zip file](BBSadaptiveCourse03Sep2023_exercises.zip) | [html](BBSadaptiveCourse03Sep2023_exercises.html).
* Solutions: follow after the course. 
* Download all files as [.zip file](BBS_adaptive.zip).








